\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-03 swmpr_manu.tex
done butchering/shortening, ready to journalify

\pkg{SWMPr} can import data into R through direct download from the CDMO or by importing local data that was previously downloaded (Table~\ref{tab:retrieve}). The IP address for the computer making the request must be registered if the first approach is used (see CDMO \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{website}).  The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Retrieval functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
\code{site\_codes}&Metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
\code{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # retrieve metadata for all sites
> site_codes()
> 
> # retrieve metadata for a single site
> site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.     
\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.  This approach is most appropriate for large, specific data requests. The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.  The \code{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \code{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path'
> 
> # this is an example path for zipped csv files
> path <- 'C:/my_path.zip'
> 
> # import, do not include file extension
> import_local(path, 'apadbwq') 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

All data retrieval functions return a \code{"swmpr"} object that includes relevant data and several attributes describing the dataset.  The data include a \code{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \code{all\_params}.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import all paramaters for the station
> # three most recent records
> exdat <- all_params('apadbwq', Max = 3, trace = F)
> exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0   13      0       NA
## 2        0      5        0     2       0  8    0    9      0       NA
## 3        0      6        0     2       0  8    0   11      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1     NA     NA       -1         
## 2         -2    NA      -1     NA     NA       -1         
## 3         -2    NA      -1     NA     NA       -1
\end{verbatim}
\end{kframe}
\end{knitrout}

The attributes for a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).  These act as metadata that are used internally by many of the package functions and are updated as the data are processed. The attributes are not visible with the raw data but can be viewed as follows.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import sample data from package
> data(apadbwq)
> dat <- apadbwq
> 
> # view all attributes of dat
> attributes(dat)
>  
> # view a single attribute of dat
> attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{data.frame}.\tabularnewline
\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{`do\_mgl'}, \code{`turb'}, etc.\tabularnewline
\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.  For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}

The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.  A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.  Available methods for the \code{"swmpr"} class are described below and can also be viewed:

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view available methods for swmpr class
> methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

A sample dataset can be downloaded for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.  These data are also included with the package as binary data files (RData) that can be loaded using the \code{data} function, as shown above.  These include \code{"swmpr"} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \code{?apacpnut}).

@@

The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).

The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.  Only applies if QAQC columns are present. \tabularnewline
\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.\tabularnewline
\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \code{comb}.\tabularnewline
\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function provided in the base installation.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # qaqc screen for a swmpr object, retain only '0'
> qaqc(dat)
> 
> # retain all data regardless of flag
> qaqc(dat, qaqc_keep = NULL)
> 
> # retain only '0' and '-1' flags
> qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view the number of observations in each QAQC flag
> qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The \code{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \code{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \code{median}, \code{min}, or \code{max}.  The entire function call, including treatment of \code{NA} values, should be passed to the \code{FUN} argument (see the examples).  The function is meant to be used after \code{qaqc} processing, although it works with a warning if QAQC columns are present.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data, filter qaqc flags
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}

A subset method added to the existing generic \code{subset} function in R is available for \code{"swmpr"} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{`>'} or \code{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \code{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> data(apaebmet)
> dat <- apaebmet
> 
> # select two parameters from dat
> subset(dat, select = c('rh', 'bp'))
> 
> # subset records greater than or equal to a date
> subset(dat, subset = '2013-01-01 0:00', operator = '>=')
> 
> # subset records within a date range
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
> 
> # subset records within a date range, select two parameters
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
+   select = c('atemp', 'totsorad'))
> 
> # remove rows/columns that do not contain data
> subset(dat, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

The \code{setstep} function formats a \code{"swmpr"} object to a continuous time series at a given time step.  This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.  The first argument of the function, \code{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \code{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \code{differ} that are greater than one half the value of \code{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \code{differ} is one half the time step.  Time steps that do not match any existing data within the limits of the \code{differ} argument are not discarded, although a corresponding data value will not be assigned.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import, qaqc removal
> data(apadbwq)
> dat <- qaqc(apadbwq)
> 
> # convert time series to two hour invervals
> # tolerance of +/- 30 minutes for matching existing data
> setstep(dat, timestep = 120, differ = 30)
> 
> # convert a nutrient time series to a continuous time series
> # then remove empty rows and columns
> data(apacpnut)
> dat_nut <- apacpnut
> dat_nut <- setstep(dat_nut, timestep = 60)
> subset(dat_nut, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  All arguments must be called explicitly since an arbitrary number of \code{"swmpr"} objects can be used as input to \code{comb}.  The function combines data by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{`union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects.  Passing \code{`intersect'} to \code{method} will create a continuous time series from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \code{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.  Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).

>>>>

\pkg{SWMPr} can import data into R through direct download from the CDMO or by importing local data that was previously downloaded (Table~\ref{tab:retrieve}). The IP address for the computer making the request must be registered if the first approach is used (see CDMO \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{website}).  The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view site metadata.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Retrieval functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
\code{site\_codes}&Get metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
\code{site\_codes\_ind}&Get metadata for all stations at a single site, wrapper to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# retrieve metadata for all sites
site_codes()

# retrieve metadata for a single site
site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.     
\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# all parameters for a station, most recent
all_params('hudscwq')

# get all parameters within a date range
all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'))

# get single parameter within a date range
all_params_dtrng('hudscwq', dtrng = c('09/01/2013', '10/01/2013'), 
  param = 'do_mgl')

# single parameter for a station, most recent
single_param('hudscwq', param = 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.  This approach is most appropriate for large data requests. The \code{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain a large number of records from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \code{import\_local} function can be used to import files directly from the zipped folder.

@@

All data retrieval functions return a \code{"swmpr"} object that includes relevant data and several attributes describing the dataset.  The data include a \code{datetimestamp} column in the timezone for a station and additional parameters for the data type (weather, nutrients, or water quality).  Corresponding QAQC columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \code{all\_params}.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0   13      0       NA
## 2        0      5        0     2       0  8    0    9      0       NA
## 3        0      6        0     2       0  8    0   11      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1     NA     NA       -1         
## 2         -2    NA      -1     NA     NA       -1         
## 3         -2    NA      -1     NA     NA       -1
\end{verbatim}
\end{kframe}
\end{knitrout}

The attributes of a \code{"swmpr"} object are descriptors that are appended to the raw data (Table~\ref{tab:attributes}).  These act as metadata that are used internally by many of the package functions and are updated as the data are processed. The attributes are not visible with the raw data but can be viewed as follows.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import sample data from package
data(apadbwq)
dat <- apadbwq

# view all attributes of dat
attributes(dat)

# view a single attribute of dat
attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{"data.frame"}.\tabularnewline
\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{'do\_mgl'}, \code{'turb'}, etc.\tabularnewline
\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually \code{"POSIXct"} unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.  For example, \code{America/Jamaica} is used for Eastern Standard Time.\end{table}

The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.  This uses the standard S3 object class system for R, such that specific methods for generic functions are developed for the object class.  A \code{"swmpr"} object also secondarily inherits methods from the \code{"data.frame"} class. Available methods for the \code{"swmpr"} class are described below and can also be viewed:

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# view available methods for swmpr class
methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

The organize functions are used to `clean' or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).

The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{f\_do\_mgl} for \code{do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station.\tabularnewline
\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.\tabularnewline
\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.\tabularnewline
\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.\tabularnewline
\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function in the base installation.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# qaqc screen for a swmpr object, retain only '0'
qaqc(dat)

# retain all data regardless of flag
qaqc(dat, qaqc_keep = NULL)

# retain only '0' and '-1' flags
qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# view the number of observations in each QAQC flag
qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

A subset method added to the existing generic \code{subset} function in R is available for \code{"swmpr"} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{'>'} or \code{'<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., \code{'2007-01-01 06:30'}).

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import data
data(apaebmet)
dat <- apaebmet

# select two parameters from dat
subset(dat, select = c('rh', 'bp'))

# subset records greater than or equal to a date
subset(dat, subset = '2013-01-01 0:00', operator = '>=')

# subset records within a date range, select two parameters
subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
  select = c('atemp', 'totsorad'))
\end{verbatim}
\end{kframe}
\end{knitrout}

The \code{setstep} function formats a \code{"swmpr"} object to a continuous time series at a given time step.  The function also has a default method making it useful for standardizing arbitrary time series to a given interval.  The first argument of the function, \code{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \code{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \code{differ} that are greater than one half the value of \code{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \code{differ} is one half the time step.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import, qaqc removal
data(apadbwq)
dat <- qaqc(apadbwq)

# convert time series to two hour invervals
# tolerance of +/- 30 minutes for matching existing data
setstep(dat, timestep = 120, differ = 30)
\end{verbatim}
\end{kframe}
\end{knitrout}

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.  A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.  As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# get nut, wq, and met data as separate objects
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

# combine nut and wq data by union
comb(swmp1, swmp2, method = 'union')

# combine nut and wq data by intersect
comb(swmp1, swmp3, method = 'intersect')

# combine nut, wq, and met data by nut time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
\end{verbatim}
\end{kframe}
\end{knitrout}
